Seniority and hierarchy configuration interaction for radicals and \\ excited states

### 2023-08-28 Manuscript/manuscript.tex
TOC
validating fabris changes

% Abstract
\begin{abstract}
Hierarchy configuration interaction (hCI) has been recently introduced as an alternative configuration interaction (CI) route combining excitation degree and seniority number, which showed to efficiently recover both dynamic and static correlations for closed-shell molecular systems [\href{https://doi.org/10.1021/acs.jpclett.2c00730}{\textit{J.~Phys.~Chem.~Lett.}~\textbf{2022}, \textit{13}, 4342}].
Here, we generalize hCI for an arbitrary reference determinant, allowing calculations for radicals and for excited states in a state-specific way.
We gauge this route against excitation-based CI (eCI) and seniority-based CI (sCI) by evaluating how different ground-state properties of radicals converge to the full CI limit.
We find that hCI outperforms or matches eCI, whereas sCI is far less accurate, in line with previous observations for closed-shell molecules.
Employing the second-order Epstein-Nesbet perturbation theory as a correction significantly accelerates the convergence of hCI and eCI.
We further explore various hCI and sCI models to calculate excitation energies of closed- and open-shell systems.
Our results underline that both the choice of the reference determinant and the set of orbitals drive the fine balance between correlation of ground and excited states.
State-specific hCI2 and higher order models perform similarly to their eCI counterparts, whereas lower orders of hCI deliver poor results.
In turn, sCI1 produces decent excitation energies for radicals, encouraging the development of related seniority-based coupled cluster methods.
%\bigskip
%\begin{center}
%        \boxed{\includegraphics[keepaspectratio,width=2in]{TOC}}
%\end{center}
%\bigskip
\end{abstract}

@@

For a reference containing unpaired electrons, an approximate CI model generally produces spin-contaminated states.
One can impose the solutions to have a well-defined spin by including additional determinants generated via higher-order hierarchies, which account for the missing spin-flip configurations. \cite{Chilkuri_2021}
The spin-contamination problem and its solution are therefore equivalent to that encountered in eCI. \cite{Maurice_1996}
\fk{Here, we have employed this procedure and considered pure spin states.
For the calculation of excitation energies, however, we have also assessed the effect of not imposing this condition.}

@@

\fk{For the state-specific models, we have further assessed the impact of not imposing spin eigenstates (the results and statistics can be found in the {\SupInf}).
This amounts to not including the appropriate spin-flip determinants lying above the hierarchy or excitation degree of choice.
The effect is minimal for both excitations of both closed- and open-shell systems, in the $\Delta$hCI1.5, $\Delta$hCI2, and $\Delta$CISD models.
The mean difference on the individual excitation energies lies between \SI{0.01}{\eV} and \SI{0.02}{\eV} for the latter two and one order of magnitude less for $\Delta$hCI1.5.
Similarly small effects are seen in the global statistics.
In turn, $\Delta$hCI1 is more affected, which would be expected given it is a low order model.
Average individual excitations of closed- and open-shell systems vary by \SI{0.26}{\eV} and \SI{0.06}{\eV}, respectively, always in the sense of improving the energies in the former case,
though not enough to cause a considerable reduction of the large absolute errors.
Overall, not constraining the CI solutions to have well-defined spin brings practical complications (as more roots have to be calculated)
and does not improve the computed excitation energies obtained with the more competitive $\Delta$CISD and $\Delta$hCI2 models.
The sCI models, discussed in the next subsection, are naturally spin eigenstates, and there is no need to enforce that.}

@@

For the full set of 50 excited states for closed-shell molecules and 19 excited states for open-shell radicals,
total energies and excitation energies obtained with
the various hCI, eCI, and sCI models considered here, in both ground-state-based and state-specific approaches;
number of determinants in the reference; saddle point order associated with the $\Delta$CSF solutions;
and the reference excitation energies and corresponding methods.
\fk{For a subset of 27 excited states for closed-shell molecules and 15 excited states for radicals,
additional total energies and excitation energies obtained with lower-order state-specific hCI and eCI models without enforcing spin pure states.}
For a subset of 16 excited states for closed-shell molecules and 6 excited states for radicals,
additional total energies and excitation energies obtained with higher-order hCI and eCI models.

>>>>

% Abstract
\begin{abstract}
Hierarchy configuration interaction (hCI) has been recently introduced as an alternative configuration interaction (CI) route combining excitation degree and seniority number, which showed to efficiently recover both dynamic and static correlations for closed-shell molecular systems [\href{https://doi.org/10.1021/acs.jpclett.2c00730}{\textit{J.~Phys.~Chem.~Lett.}~\textbf{2022}, \textit{13}, 4342}].
Here, we generalize hCI for an arbitrary reference determinant, allowing calculations for radicals and for excited states in a state-specific way.
We gauge this route against excitation-based CI (eCI) and seniority-based CI (sCI) by evaluating how different ground-state properties of radicals converge to the full CI limit.
We find that hCI outperforms or matches eCI, whereas sCI is far less accurate, in line with previous observations for closed-shell molecules.
Employing the second-order Epstein-Nesbet perturbation theory as a correction significantly accelerates the convergence of hCI and eCI.
We further explore various hCI and sCI models to calculate excitation energies of closed- and open-shell systems.
Our results underline that both the choice of the reference determinant and the set of orbitals drive the fine balance between correlation of ground and excited states.
State-specific hCI2 and higher order models perform similarly to their eCI counterparts, whereas lower orders of hCI deliver poor results.
In turn, sCI1 produces decent excitation energies for radicals, encouraging the development of related seniority-based coupled cluster methods.
\bigskip
\begin{center}
        \boxed{\includegraphics[keepaspectratio,width=3.25in]{TOC}}
\end{center}
\bigskip
\end{abstract}

@@

For a reference containing unpaired electrons, an approximate CI model generally produces spin-contaminated states.
One can impose the solutions to have a well-defined spin by including additional determinants generated via higher-order hierarchies, which account for the missing spin-flip configurations. \cite{Chilkuri_2021}
The spin-contamination problem and its solution are therefore equivalent to that encountered in eCI. \cite{Maurice_1996}
Here, we have employed this procedure and considered pure spin states.
For the calculation of excitation energies, however, we have also assessed the effect of not imposing this condition.

@@

For the state-specific models, we have further assessed the impact of not imposing spin eigenstates (the results and statistics can be found in the {\SupInf}).
This amounts to not including the appropriate spin-flip determinants lying above the hierarchy or excitation degree of choice.
The effect is minimal for the excitation energies of both closed- and open-shell systems, in the $\Delta$hCI1.5, $\Delta$hCI2, and $\Delta$CISD models.
The mean difference on the individual excitation energies lies between \SI{0.01}{\eV} and \SI{0.02}{\eV} for the latter two and one order of magnitude less for $\Delta$hCI1.5.
Similarly, small effects are seen in the global statistics.
In turn, $\Delta$hCI1 is more affected, which would be expected given it is a low order model.
Average individual excitations of closed- and open-shell systems vary by \SI{0.26}{\eV} and \SI{0.06}{\eV}, respectively, always in the sense of improving the energies in the former case,
though not enough to cause a considerable reduction of the large absolute errors.
Overall, not constraining the CI solutions to have well-defined spin brings practical complications (as more roots have to be calculated)
and does not improve the computed excitation energies obtained with the more competitive $\Delta$CISD and $\Delta$hCI2 models.
The sCI models, discussed in the next subsection, are naturally spin eigenstates, and there is no need to enforce that.

@@

For the full set of 50 excited states for closed-shell molecules and 19 excited states for open-shell radicals,
total energies and excitation energies obtained with
the various hCI, eCI, and sCI models considered here, in both ground-state-based and state-specific approaches;
number of determinants in the reference; saddle point order associated with the $\Delta$CSF solutions;
and the reference excitation energies and corresponding methods.
For a subset of 27 excited states for closed-shell molecules and 15 excited states for radicals,
additional total energies and excitation energies obtained with low-order state-specific hCI and eCI models without enforcing spin pure states.
For a subset of 16 excited states for closed-shell molecules and 6 excited states for radicals,
additional total energies and excitation energies obtained with higher-order hCI and eCI models.
